TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-06-11 paper/draft.tex
adjust paragraph spacing

This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the \apogee\ and \gaia\ surveys, as well as previous studies of M dwarfs in \apogee.
Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present and upcoming stellar and exoplanetary surveys.

@@

Need to talk about \gaia\ data \& what we use this dataset for...

@@

The \emph{training step} consists of optimizing the likelihood function (Equation \ref{eqn:likelihood}) for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) given fixed label vector (\emph{$\ell_n$}) constructed from the reference labels, and the \emph{test step} consists of optimizing the likelihood function for the labels at fixed $\theta_{\lambda}$ and $s_{\lambda}$ obtained in the training step (see \citealt{Ness:2015} for further description). 
In the training step, the regression is designed to predict spectral pixels
given labels, by learning zeroth, first, and second derivatives of the data with respect to
the labels. In the test step, the regression is designed to predict labels given the spectral
derivatives.

@@

We note that surface gravity is not included as a training label. The reason for this is that for main-sequence M dwarfs, the parameter is almost entirely redundant with metallicity. Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe after arriving at the zero-age main sequence. Hence perfect knowledge of abundances and $\teff$ for an M dwarf should uniquely determine its surface gravity, position on a color-magnitude diagram, and overall luminosity. While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important compared to other uncertainties in extreme cases (e.g., Carbon stars).

@@

\color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me? I am imagining a figure that shows a scatter plot of T vs RMS residual, colored by metallicity. Or something like that.}\color{black} \\

\color{gcolor}{HOGG: Adam's comment: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\

@@

Figure \ref{fig:isochrone_cmd}

@@

Given that physics-based spectral models of M dwarfs have issues, one of the possible
future values of the data-driven model shown here is that it is highly interpretable:
It contains within it first and second derivatives of the spectral expectation with
respect to the atmospheric parameters.
We show some of these derivatives in [BIRKY WHAT FIGURE] and deliver relevant data in
Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}
These Tables summarize spectral features in the \apogee\ bandpass
that are found to be strong temperature and metallicity indicators.
In the long run, this is the primary value of data-driven models for astronomy:
To provide physical insights that drive physical understandings.
It is our hope that \thecannon, and models like it, will, in the end, lead to
new and improved physical models, which will, in turn, put \thecannon\ out of business.

>>>>

This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the \apogee\ and \gaia\ surveys, as well as previous studies of M dwarfs in \apogee.
Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present and upcoming stellar and exoplanetary surveys. \\

@@

Need to talk about \gaia\ data \& what we use this dataset for... \\

@@

The \emph{training step} consists of optimizing the likelihood function (Equation \ref{eqn:likelihood}) for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) given fixed label vector (\emph{$\ell_n$}) constructed from the reference labels, and the \emph{test step} consists of optimizing the likelihood function for the labels at fixed $\theta_{\lambda}$ and $s_{\lambda}$ obtained in the training step (see \citealt{Ness:2015} for further description). 
In the training step, the regression is designed to predict spectral pixels
given labels, by learning zeroth, first, and second derivatives of the data with respect to
the labels. In the test step, the regression is designed to predict labels given the spectral
derivatives. \\

@@

We note that surface gravity is not included as a training label. The reason for this is that for main-sequence M dwarfs, the parameter is almost entirely redundant with metallicity. Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe after arriving at the zero-age main sequence. Hence perfect knowledge of abundances and $\teff$ for an M dwarf should uniquely determine its surface gravity, position on a color-magnitude diagram, and overall luminosity. While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important compared to other uncertainties in extreme cases (e.g., Carbon stars).  \\

@@

% \color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me? I am imagining a figure that shows a scatter plot of T vs RMS residual, colored by metallicity. Or something like that.}\color{black} \\

% \color{gcolor}{HOGG: Adam's comment: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\

@@

Figure \ref{fig:isochrone_cmd} \\

@@

Given that physics-based spectral models of M dwarfs have issues, one of the possible
future values of the data-driven model shown here is that it is highly interpretable:
It contains within it first and second derivatives of the spectral expectation with
respect to the atmospheric parameters.
We show some of these derivatives in [BIRKY WHAT FIGURE] and deliver relevant data in
Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}
These Tables summarize spectral features in the \apogee\ bandpass
that are found to be strong temperature and metallicity indicators.
In the long run, this is the primary value of data-driven models for astronomy:
To provide physical insights that drive physical understandings.
It is our hope that \thecannon, and models like it, will, in the end, lead to
new and improved physical models, which will, in turn, put \thecannon\ out of business. \\
